---
title: "Demonstration of the disclim Package for Implementing Climate Driven, Trait-Based
  Models of Disease Transmission"
author: "Michael C. Wimberly"
date: "2024-02-27"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup

The tidyverse suite of packages is used for data manipulation and graphing, and devtools is required for the `load_all()` function. Right now `load_all()`is used to read the disclim functions from a local folder, but this should eventually be changed to install the package from GitHub and then load normally with `library()`.

```{r loadpackages, warning=FALSE, message=FALSE}
library(tidyverse)
library(devtools)
mypath <- "C:/Users/wimb0002/OneDrive - University of Oklahoma/Work/disclim"
load_all(mypath)
```

Next, files containing the parameters are loaded. There are two sets of parameters from Miazgawicz et al. (2020) for malaria transmission by *Anopheles stephensi*, which correpond to the "estimated' and "lifetime" models. There is also a set of parameters from Mordecai et al. (2017) for dengue transmission by *Aedes aegypti*. Parameters are stored in CSV files that are currently being read in using the base R function `read.csv()`. In the future, it would be nice to have a package function that would read in the parameter files and check them for consistency.

```{r readparams}
mal_est_par <- read.csv(file.path(mypath, "Anstephensi_mal_est_Miazgowicz20.csv"))
den_aegypti_par <- read.csv(file.path(mypath, "Aeaegypti_dengue_Mordecai17.csv"))
```

## Temperature-trait relationships

Each parameter table is a data frame with columns for 1) the variables in the model, 2) the forms of the corresponding temperature trait curves, and 3) values of the rate constant (rc) minimum temperature (tmin) and maximum temperature (tmax) parameters for each curve. The `getpar()` function retrieves parameters for a given variable from a parater table.

```{r printparams}
mal_est_par
getpar("a", "rc", mal_est_par)
getpar("a", "tmin", mal_est_par)
getpar("a", "tmax", mal_est_par)
```

The `tempcurve()` function is used to calculate the trait values as a function of temperature for each variable. The function is vectorized, so a range of temperature can be provided.

```{r tempcurve1}
Temp <- seq(1, 50, 0.01)
a <- tempcurve(Temp, "a", mal_est_par)
ggplot() +
  geom_line(aes(x = Temp, y = a))

```

This process can be repeated for all the trait variables in a particular model.

```{r tempcurve2}
lf <- tempcurve(Temp, "lf", mal_est_par)
EFD <- tempcurve(Temp, "EFD", mal_est_par)
PDR <- tempcurve(Temp, "PDR", mal_est_par)
bc <- tempcurve(Temp, "bc", mal_est_par)
pEA <- tempcurve(Temp, "pEA", mal_est_par)
MDR <- tempcurve(Temp, "MDR", mal_est_par)

traitvals <- c(a, lf, EFD, PDR, bc, pEA, MDR)
tempvals <- rep(Temp, 7)
traitnames <- c("a", "lf", "EFD", "PDR", "bc", "pEA", "MDR")
trait <- rep(traitnames, each = length(Temp))
traitdf <- data.frame(tempvals, traitvals, trait)

ggplot(data = traitdf) +
  geom_line(aes(x = tempvals, y = traitvals)) +
  facet_wrap(~ trait, ncol = 4, scales = "free_y") +
  labs(title = 'Malaria trait curves from Miazgowicz et al. (2020) "estimated" model',
       x = "Temperature",
       y = "Trait Values")
```
# Vectorial capacity

The `vc_rossmac()` function implements the temperature-dependent vectorial capacity (VC) equation derived from the Ross-Madonald malaria transmission model. Arguments include a vector of temperature values and a parameter table. The `bc` argument is set fo `TRUE` by default, indicating that a single `bc` trait variable is used for transmission. The argument should be set to `FALSE` for models where there are separate `b` and `c` trait variables, as is the case with the Mordecai et al. (2017) dengue transmission models. There is a separate `vc_life()` function for calculating VC based on the "lifetime" model from Miazgowicz et al. (2020).


```{r vc}
mal_est_m <- m_trait(Temp, 
                     lf = lf, 
                     EFD = EFD, 
                     pEA = pEA, 
                     MDR = MDR)

mal_est_vc <- vc_rossmac(Temp, 
                         M = mal_est_m,
                         a = a,
                         lf = lf,
                         PDR = PDR,
                         bc = bc,
                         bcvar = TRUE)

mal_vals <- c(mal_est_m, mal_est_vc)
tempvals <- rep(Temp, 2)
modelnames <- c("M", "VC")
model <- rep(modelnames, each = length(Temp))
mal_df <- data.frame(tempvals, mal_vals, model)
ggplot(data = mal_df) +
  geom_line(aes(x = tempvals, y = mal_vals, color = model)) +
  facet_wrap(~ model, scales = "free_y", ncol = 1) +
  labs(y = "Model Output", x = "Temperature")
```

The vectorial capacity equations incorporate trait-based estimates of the mosquito abundance parameter (M). These values can also be extracted nad plotted as a function of temperature.

```{r tempdurve3}
a <- tempcurve(Temp, "a", den_aegypti_par)
lf <- tempcurve(Temp, "lf", den_aegypti_par)
EFD <- tempcurve(Temp, "EFD", den_aegypti_par)
PDR <- tempcurve(Temp, "PDR", den_aegypti_par)
b <- tempcurve(Temp, "b", den_aegypti_par)
c <- tempcurve(Temp, "c", den_aegypti_par)
pEA <- tempcurve(Temp, "pEA", den_aegypti_par)
MDR <- tempcurve(Temp, "MDR", den_aegypti_par)

traitvals <- c(a, lf, EFD, PDR, b, c, pEA, MDR)
tempvals <- rep(Temp, 8)
traitnames <- c("a", "lf", "EFD", "PDR", "b", "c", "pEA", "MDR")
trait <- rep(traitnames, each = length(Temp))
traitdf <- data.frame(tempvals, traitvals, trait)

ggplot(data = traitdf) +
  geom_line(aes(x = tempvals, y = traitvals)) +
  facet_wrap(~ trait, ncol = 4, scales = "free_y") +
  labs(title = 'Dengue trait curves from Mordecai et al. (2017) Ae. aegypti model',
       x = "Temperature",
       y = "Trait Values")

```



```{r vc2}
den_est_m <- m_trait(Temp, 
                     lf = lf, 
                     EFD = EFD, 
                     pEA = pEA, 
                     MDR = MDR)

den_est_vc <- vc_rossmac(Temp, 
                         M = mal_est_m,
                         a = a,
                         lf = lf,
                         PDR = PDR,
                         b = b,
                         c = c,
                         bcvar = FALSE)

den_vals <- c(den_est_m, den_est_vc)
tempvals <- rep(Temp, 2)
modelnames <- c("M", "VC")
model <- rep(modelnames, each = length(Temp))
den_df <- data.frame(tempvals, den_vals, model)
ggplot(data = den_df) +
  geom_line(aes(x = tempvals, y = den_vals, color = model)) +
  facet_wrap(~ model, scales = "free_y", ncol = 1) +
  labs(y = "Model Output", x = "Temperature")
```
